agdarsec -- Total Parser Combinators

### 2017-10-26 doc/paper/main.tex
[ doc ] more about the total parser combinators

\DeclareUnicodeCharacter{8759}{:\,:}

@@

The first kind of combinator corresponds to operations on
sets which are lifted to indexed sets by silently propagating
the index. It may be a bit surprising to see infix notations
(in Agda underscores correspond to positions in which arguments
are to be inserted) for functions taking three arguments but
they are only meant to be partially applied.

\sagrab{Relation/Unary/Indexed}{arrow}
\agrab{Relation/Unary/Indexed}{product}

@@

We can already see the benefits of using these aliases. For instance
the fairly compact expression \AF{[} (\AB{P} \AF{⊗} \AB{Q}) \AF{⟶} \AB{R} \AF{]}
corresponds to the more verbose type
\AS{∀} \{\AB{n}\} \AS{→} (\AB{P} \AB{n} \AF{×} \AB{Q} \AB{n}) \AS{→} \AB{R} \AB{n}.

@@

\remark{Counit} The careful reader will have noticed
that this is not quite the extract we would expect from a comonad:
for a counit, we would need a natural transformation between
\BOX{} \AB{A} and \AB{A} i.e. a function of type
\AF{[} \BOX{} \AB{A} \AF{⟶} \AB{A} \AF{]}. We will not be able
to define such a function: \BOX{} \AB{A} \AN{0} is isomorphic to
the unit type so we would have to generate an \AB{A} \AN{0} out
of thin air.

@@

\remark{Implicit Field} Just like the arguments to a function can be
implicit, so can a record's fields. The user can them leave them out
when building a value and they will be filled in by unification.

@@

@@

@@

>>>>

\DeclareUnicodeCharacter{8759}{:\,:}
\DeclareUnicodeCharacter{954}{$\kappa$}

@@

The first kind of combinators corresponds to operations on
sets which are lifted to indexed sets by silently propagating
the index. We only show the ones we will use in this paper:
the pointwise arrow and product types and the constant function.
It may be a bit surprising to see infix notations (in Agda
underscores correspond to positions in which arguments are to
be inserted) for functions taking three arguments but they are
only meant to be partially applied.

\sagrab{Relation/Unary/Indexed}{arrow}
\agrab{Relation/Unary/Indexed}{product}
\agrab{Relation/Unary/Indexed}{constant}

@@

We can already see the benefits of using these aliases. For instance
the fairly compact expression \AF{[} (\AF{κ} \AB{P} \AF{⊗} \AB{Q}) \AF{⟶} \AB{R} \AF{]}
corresponds to the more verbose type
\AS{∀} \{\AB{n}\} \AS{→} (\AB{P} \AF{×} \AB{Q} \AB{n}) \AS{→} \AB{R} \AB{n}.

@@

\remark{Counit} The careful reader will have noticed
that this is not quite the extract we would expect from a comonad:
for a counit, we would need a natural transformation between
\BOX{} \AB{A} and \AB{A} i.e. a function of type
\AF{[} \BOX{} \AB{A} \AF{⟶} \AB{A} \AF{]}. We will not be able
to define such a function: \BOX{} \AB{A} \AN{0} is isomorphic to
the unit type so we would have to generate an \AB{A} \AN{0} out
of thin air. The types \AB{A} for which \BOX{} has a counit are
interesting in their own right: they are inhabited at every single
index as demonstrated by \AF{fix} later on.

@@

\remark{Implicit Field} Just like the arguments to a function can be
implicit, so can a record's fields. The user can then leave them out
when building a value and they will be filled in by unification.

@@

Now that we have a precise definition of \AR{Parser}s, we can start
building our library of combinators. Our first example \AF{anyChar}
can be defined by copattern-matching and then case analysis on the
input string: if it is empty then the list of \AR{Success}es is also
empty, otherwise it contains exactly one element which corresponds to
the head of the input string and its tail as leftovers.

@@

Unsurprisingly \AF{guardM} is still a valid higher-order combinator:
filtering out results which do not agree with a predicate is absolutely
compatible with the consumption constraint we have drawn. To implement
\AF{guardM} we can once more reuse existing library functions such
\AF{gfilter} which turns a \mbox{\AD{List} \AB{A}} into a \mbox{\AD{List} \AB{B}}
provided a predicate \mbox{\AB{A} \AS{→} \AD{Maybe} \AB{B}}.

@@

Demonstrating that \AR{Parser} is a functor goes along the same
lines: using \AD{List}'s and \AR{Success}'s \AF{map}s. Similarly,
we can prove that it is an \AF{Alternative}: failing corresponds
to returning the empty list no matter what whilst disjunction is
implemented using concatenation.

\agrab{Text/Parser/Combinators}{fmap}\vspace*{-2em}
\agrab{Text/Parser/Combinators}{fail}\vspace*{-2em}
\agrab{Text/Parser/Combinators}{disjunction}

So far the types we have ascribed to our combinators are exactly the
same as the ones one would find in any other parsec library. In none
of the previous combinators do we run a second parser on the leftovers
of a first one. All we do is either manipulate or combine the results
of one or more parsers run in parallel, potentially discarding some of
these results on the way.

However when we run a parser \emph{after} some of the input has already
been consumed, we could safely perform a \emph{guarded} call. This being
made explicit would be useful when using \AF{fix} to define a parser for
a recursive grammar. Luckily \AR{Parser} is, by definition, a
downwards-closed type. This means that we may use very precise types marking
all the guarded positions with \BOX{}; if the user doesn't need that extra
power she can very easily bypass the \BOX{} annotations by using \AF{box}:

\agrab{Text/Parser/Combinators}{box}

The most basic example we can give of such an annotation is probably the
definition of a conjunction combinator \AF{\_<\&>\_} taking two parsers,
running sequentially and returning a pair of their results. The second
parser is given the type \mbox{\BOX{} \AR{Parser} \AB{B}} instead of
\mbox{\AR{Parser} \AB{B}} which we would expect to find in other parsec libraries.

\agrab{Text/Parser/Combinators}{conjunction}

We can immediately use all of these newly-defined combinator to give
a safe, total definition of \AF{some} which takes a parser for \AB{A}
and returns a parser for non-empty lists of \AB{A}s. It is defined as
a fixpoint and proceeds as follows: it either combines a head and a
non-empty tail using \AF{$\_::^{+}\_$} or returns a singleton list.

\agrab{Text/Parser/Combinators}{badsome}

\remark{Inefficiency} Unfortunately this definition is inefficient.
Indeed, in the base case \mbox{\AF{some} \AB{p}} is going to run the parser
\AB{p} twice: once in the first branch before realising that \AB{rec}
fails and once again in the second branch. Compare this to the Haskell
version (after inlining \AB{many}):

\shgrab{some2}

This inefficiency can be fixed by using a notion of conjunction
\AF{\_<\&?>\_} which allows the second parser to fail whilst still
producing a successful run. The parser \mbox{\AF{some} \AB{p}} can then be
defined as the conjunction of \AB{p} and a potentially failing
recursive call. The function \AF{cons} (whose definition we omit here)
either uses \AF{$\_::^{+}\_$} to put the head and the non-empty
tail together or returns a singleton list if the recursive call
failed.

\agrab{Text/Parser/Combinators}{conjunction2}
\agrab{Text/Parser/Combinators}{goodsome}

\remark{Non-compositionality} The higher-order parser \AF{some}
expects a \emph{fully} defined parser as an argument. This makes
it impossible to use it as one of the building blocks of a larger,
recursive parser. Ideally we would rather have a combinator of
type \mbox{\AF{[} \AF{Parser} \AB{A} \AF{⟶} \AF{Parser} (\AF{List$^+$} \AB{A}) \AF{]}}.
This will be adressed in the next subsection.
\medskip{}

As the previous code snippet shows, \AF{\_<\&?>\_} is defined in
terms of a more fundamental notion \AF{\_\&?>>=\_} which is a
combinator analogous to a monad's \textit{bind}. On top of running
two parsers sequentially (with the second one being chosen based on
the result of running the first), it allows the second one to fail
and returns both results.

\agrab{Text/Parser/Combinators}{mbind}

This crucial definition makes it possible to port a lot of the Haskell
definitions where one would use a parser which does not use any of its
input. And this is possible without incurring any additional cost as
the optimised version of \AF{some} showed.

\subsection{Left and Right Chains}

The pattern used in the solution presented in section~\ref{sec:haskellproblem}
can be abstracted with the notion of an (heterogeneous) left chain which takes
a parser for a seed, one for a constructor, and one for and argument. The crucial
thing is to make sure \emph{not} to use the parser one is currently defining as
the seed.

\shgrab{hchainl}

We naturally want to include a safe variant of this combinator in our
library. However this definition relies on the ability to simply use
\textit{return} in case it's not possible to parse an constructor
additional constructor and argument and that is something we simply
don't have access to.
